---
title: "final"
output: pdf_document
date: "`r Sys.Date()`"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(msa)
library(castor)
library(phangorn)
library(phytools)
library(ape)
```

```{r}
#Import the aligned sequence using Clustal Omega
#module load clustal-omega/1.2.4
#clustalo -i cetacean_final.fasta -o cetacean_alignment.fasta --threads=4

alignedSeqs = readDNAStringSet("cetaceans_aligned.fasta")

#Use phangorn to compare DNA substitution models
forPhang = phyDat(as.matrix(alignedSeqs), type = "DNA")
mt = modelTest(forPhang) # run a model test for all models(20+)
write.table(mt, "modelTest_results.txt", quote=FALSE, sep="\t",row.names=FALSE, col.names=TRUE) #Write a table for results of mt test

#D = dist.dna(dna, model="GTR", gamma=TRUE)
```

Partitioning the Ceteceans file
```{r}
#Prepwork for partitioning
# Convert the alignment to a DNAMultipleAlignment object
concat.aln = DNAMultipleAlignment(alignedSeqs)
# Write the alignment to a Phylip and Nexus file for use in RAxML and MrBayes
write.phylip(concat.aln, "cetaceansConcat.phy")
write.nexus.data(as.DNAbin(concat.aln), file="cetaceansConcat.nex")

#Create the Partition File for RAxML
# Define the start and end positions for each gene
#my.starts = c(1, 850, 1759, 3100, 4954, 5920, 6740)
#my.ends = c(849, 1758, 3099, 4953, 5919, 6738, 7392)
my.starts = c(1, 910, 1819, 3160, 5014, 5980, 6799)
my.ends = c(909, 1818, 3159, 5013, 5979, 6798, 7452)
# Define the gene names
my.genes = c("AQP6", "FBXO2", "FGG", "FOXP1", "HOXA13", "HOXA9", "HOXB7")
# Create formatted information for each gene
info1 = paste(my.starts, my.ends, sep="-")
info2 = paste(my.genes, info1, sep="=")
info3 = paste("DNA,", info2, sep=" ")
# Write the partition information to a file
write.table(info3, file="cetaceans-Raxml-part.txt", quote=FALSE, row.names=FALSE, col.names=FALSE)
```
SLURM script for running RAxML
```{r, engine='bash', eval=FALSE}
#!/bin/bash
#SBATCH --time=4:00:00
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --mem=32GB
#SBATCH --partition=Centaurus
#SBATCH --mail-user=mserkin@uncc.edu
#SBATCH --mail-type=ALL

module load raxml

raxmlHPC -s cetaceansConcat.phy \
    -n cetaceansboot \
    -f a \
    -q cetaceans-Raxml-part.txt \
    -M \
    -m GTRGAMMAI \
    -N 1000 \
    -o Hippo \
    -p 9503 -x 9004 \
    >cetaceans_raxml_log.txt
```
Explaining the parameters for the slurm script:
-s      the input alignment file
-n      the name of the output files to be created 
-f a    specifies a rapid bootstrap algorithm
-m      the model we use (we have only used GTRGAMMA)
-N      the number of bootstrap replicates to be performed
-o      the outgroup for rooting the tree
-p -x   random number seeds for the bootstrap and tree searching algorithms

Plot the raxml-tree
```{r raxml, fig.dim=c(8,8), fig.keep="last"}
my.tree = read.tree("RAxML_bestTree.cetaceansboot")
my.boot = read.tree("RAxML_bootstrap.cetaceansboot")
x = plotBS(my.tree, my.boot, cex=0.7, main="RAxML Tree")
plot(x)
nodelabels(x$node.label, cex=0.7)

print(x$node.label)
```
Continual ASR for Brain Size
```{r asr-all-equal, fig.dim=c(8,8)}
# Load your tree
my.tree <- read.tree("RAxML_bestTree.cetaceansboot")
plot(my.tree, main="Cetaceans RAxML Tree")

# Define species names and log10 brain mass values
species_names = c(
  "Bbor", "Bmus49099", "Bphy_baits", "Cmar59905", "DdelSJR",
  "Gmac39091", "Ggri", "Ksim", "Lalb", "Lobs",
  "Lvex", "Meur", "Oorc", "Pmac", "Pcra123188",
  "Scly1726", "Scoe", "Sbre116871", "Ttru", "Hippo"
)

log10_brain_mass = c(
  3.693 , 3.827, 3.851, 3.104, 2.85, 3.43, 3.33, 2.801, 3.05, 2.94,
  2.76, 3.23, 3.82, 3.89, 3.59, 2.81, 2.94, 3.16, 3.19, 2.77
)

names(log10_brain_mass) <- species_names

# Ensure the tree tip labels and trait data align
m <- match(my.tree$tip.label, names(log10_brain_mass))
numeric_scores <- log10_brain_mass[m]

# Use contMap to visualize the continuous trait with colors
contMap_result <- contMap(my.tree, numeric_scores, plot=TRUE, fsize=0.8)

# Customize colors for better readability
contMap_result <- setMap(contMap_result, colors=rainbow(100, start=0, end=0.85))

# Plot again with the updated color gradient
plot(contMap_result)
```

Discrete ASR Graph for Brain Size (not used, just helped visually)
```{r}
# Load your tree
my.tree = read.tree("RAxML_bestTree.cetaceansboot")
plot(my.tree, main = "Cetaceans RAxML Tree")

# Species names and log10 brain mass data
species_names = c(
  "Bbor", "Bmus49099", "Bphy_baits", "Cmar59905", "DdelSJR",
  "Gmac39091", "Ggri", "Ksim", "Lalb", "Lobs",
  "Lvex", "Meur", "Oorc", "Pmac", "Pcra123188",
  "Scly1726", "Scoe", "Sbre116871", "Ttru", "Hippo"
)

log10_brain_mass = c(
  3.693 , 3.827, 3.851, 3.104, 2.85, 3.43, 3.33, 2.801, 3.05, 2.94,
  2.76, 3.23, 3.82, 3.89, 3.59, 2.81, 2.94, 3.16, 3.19, 2.77
)

# Assign species names as names for brain mass
names(log10_brain_mass) = species_names

# Align log10_brain_mass with tree tip labels
m = match(my.tree$tip.label, names(log10_brain_mass))  # Match tree tip labels to brain mass names
log10_brain_mass = log10_brain_mass[m]  # Rearrange brain mass in the correct order

# Recalculate quantiles for discrete states
quantiles = quantile(log10_brain_mass, probs = c(0, 0.25, 0.5, 0.75, 1))
print(quantiles)  # Show the quantile breaks

# Assign discrete scores (states) based on quantiles
disc_scores = cut(
  log10_brain_mass,
  breaks = quantiles,
  labels = c(1, 2, 3, 4),  # State 1, 2, 3, 4
  include.lowest = TRUE
)

# Ensure names for disc_scores align with the tree tip labels
names(disc_scores) = my.tree$tip.label

# Verify alignment of data
print(data.frame(
  TipLabel = my.tree$tip.label,
  BrainMass = log10_brain_mass,
  DiscScore = disc_scores
))

# Ensure numeric states for ancestral state reconstruction
numeric_states = as.numeric(as.factor(disc_scores))

# Perform ancestral state reconstruction with equal transition costs
mp_equal = asr_max_parsimony(
  tree = my.tree,
  tip_states = numeric_states,
  Nstates = length(unique(numeric_states)),
  transition_costs = "all_equal"
)

# Create a pie matrix for visualization
pie_matrix = matrix(mp_equal$ancestral_likelihoods, ncol = length(unique(numeric_states)))
rownames(pie_matrix) = seq(length(my.tree$tip.label) + 1, length(my.tree$tip.label) + nrow(pie_matrix))
colnames(pie_matrix) = paste("State", 1:length(unique(numeric_states)))

# Plot the tree with pie charts for each node
state_labels = c(
  "State 1: < 2.92",
  "State 2: 2.92 - 3.39",
  "State 3: 3.39 - 3.67",
  "State 4: > 3.67"
)

# Plot tree and visualize results
plotTree(my.tree, offset = 0.5, main = "Ancestral Reconstruction with Fixed Data")

# Add pie charts to the tips
tiplabels(
  pie = to.matrix(disc_scores, sort(unique(disc_scores))),
  piecol = c("navy", "skyblue", "green", "yellow"),
  cex = 0.3
)

# Add pie charts to the nodes
nodelabels(
  node = as.numeric(rownames(pie_matrix)),
  pie = pie_matrix,
  piecol = c("navy", "skyblue", "green", "yellow"),
  cex = 0.5
)

# Add legend
legend(
  "bottomright",
  legend = state_labels,
  fill = c("navy", "skyblue", "green", "yellow"),
  bty = "n"
)

```


LANGAUGE COMPLEXITY SUBSET:
```{r}
#ALIGN THE DATA
#Import the aligned sequence using Clustal Omega
#module load clustal-omega/1.2.4
#clustalo -i language_ceteceans.fasta -o language_ceteceans_aligned.fasta --threads=4

alignedSeqs = readDNAStringSet("language_ceteceans_aligned.fasta")

#MODEL TEST
#Use phangorn to compare DNA substitution models
forPhang = phyDat(as.matrix(alignedSeqs), type = "DNA")
mt = modelTest(forPhang) # run a model test for all models(20+)
write.table(mt, "modelTest_results.txt", quote=FALSE, sep="\t",row.names=FALSE, col.names=TRUE) #Write a table for results of mt test

#D = dist.dna(dna, model="GTR", gamma=TRUE)
```

Partitioning the Langauge subset file
```{r}
#Prepwork for partitioning
# Convert the alignment to a DNAMultipleAlignment object
concat.aln = DNAMultipleAlignment(alignedSeqs)
# Write the alignment to a Phylip and Nexus file for use in RAxML and MrBayes
write.phylip(concat.aln, "languageConcat.phy")
write.nexus.data(as.DNAbin(concat.aln), file="languageConcat.nex")

#Create the Partition File for RAxML
# Define the start and end positions for each gene
#my.starts = c(1, 850, 1759, 3100, 4954, 5920, 6740)
#my.ends = c(849, 1758, 3099, 4953, 5919, 6738, 7392)
my.starts = c(1, 906, 1815, 3156, 5010, 5976, 6795)
my.ends = c(905, 1814, 3155, 5009, 5975, 6794, 7448)
# Define the gene names
my.genes = c("AQP6", "FBXO2", "FGG", "FOXP1", "HOXA13", "HOXA9", "HOXB7")
# Create formatted information for each gene
info1 = paste(my.starts, my.ends, sep="-")
info2 = paste(my.genes, info1, sep="=")
info3 = paste("DNA,", info2, sep=" ")
# Write the partition information to a file
write.table(info3, file="language-Raxml-part.txt", quote=FALSE, row.names=FALSE, col.names=FALSE)
```
SLURM script for running RAxML
```{r, engine='bash', eval=FALSE}
#!/bin/bash
#SBATCH --time=4:00:00
#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --mem=32GB
#SBATCH --partition=Centaurus
#SBATCH --mail-user=mserkin@uncc.edu
#SBATCH --mail-type=ALL

module load raxml

raxmlHPC -s languageConcat.phy \
    -n languageboot \
    -f a \
    -q language-Raxml-part.txt \
    -M \
    -m GTRGAMMAI \
    -N 1000 \
    -o Hippo \
    -p 9503 -x 9004 \
    >language_raxml_log.txt
```

RAXML TREE Subset
```{r fig.dim=c(8,8), fig.keep="last"}
my.tree = read.tree("RAxML_bestTree.languageboot")
my.boot = read.tree("RAxML_bootstrap.languageboot")
x = plotBS(my.tree, my.boot, cex=0.8, main="RAxML Tree")
plot(x)
nodelabels(x$node.label, cex=0.7)
```

Discrete ASR for Language Complexity
```{r}
my.tree = read.tree("RAxML_bestTree.languageboot")

species_names = c(
  "DdelSJR", "Gmac39091",
  "Ggri",
  "Oorc", "Pmac", "Pcra123188",
  "Scoe",
  "Ttru", "Hippo"
)

# Create a vector for language complexity
language_complexity = c( 8, 8, 7, 10, 10, 8, 9, 10, 6)

names(language_complexity) = species_names
m = match(my.tree$tip.label, names(language_complexity))
language_complexity = language_complexity[m]

# Assign discrete scores based on complexity
my.complex = language_complexity
my.complex[which(language_complexity == 6)] = 1  
my.complex[which(language_complexity == 7)] = 2  
my.complex[which(language_complexity == 8)] = 3 
my.complex[which(language_complexity == 9)] = 4  
my.complex[which(language_complexity == 10)] = 5  

#Make sure these are in the same order as the tip labels
m = match(my.tree$tip.label, names(my.complex)) 
my.complex = my.complex[m]

Q = matrix(data=c(
  0,  2,  4,  6,  8,    
  2,  0,  2,  4,  6,   
  4,  2,  0,  2,  4,     
  6,  4,  2,  0,  2,      
  8,  6,  4,  2,  0    
), nrow=5, byrow=TRUE, dimnames=list(c("1", "2", "3", "4", "5"), c("1", "2", "3", "4", "5")))

my.complex = my.complex[m]

# Check if the names match the tip labels in the tree
tree_tip_labels = my.tree$tip.label

# Ensure numeric states for ASR
numeric_states = as.numeric(as.factor(my.complex))

mp_custom = asr_max_parsimony(tree=my.tree, tip_states=numeric_states, Nstates=5, transition_costs=Q)


pie.matrix = matrix(mp_custom$ancestral_likelihoods, ncol=5)
rownames(pie.matrix) = seq(10,17) #(number of taxa+1, number of taxa+number of rows)
colnames(pie.matrix) = c("1", "2", "3", "4", "5")
plotTree(my.tree, offset=0.5, main="Ancestral Reconstruction with Custom Transition Costs")

tiplabels(pie = to.matrix(numeric_states, sort(unique(numeric_states))), 
          piecol=c("yellow","hotpink","seagreen",
                   "skyblue","navy"), 
          cex=0.3)
nodelabels(node = as.numeric(rownames(pie.matrix)), 
           pie=pie.matrix, piecol=c("yellow","hotpink","seagreen",
                                     "skyblue","navy"),
           cex=0.5)

legend("bottomright", 
       levels(as.factor(numeric_states)),
       fill=c("yellow","hotpink","seagreen",
              "skyblue","navy"),
       bty="n", 
       title = "Language
      Complexity")
```

Discrete ASR for Brain Size
```{r}
my.tree = read.tree("RAxML_bestTree.languageboot")

species_names = c(
  "DdelSJR", "Gmac39091",
  "Ggri",
  "Oorc", "Pmac", "Pcra123188",
  "Scoe",
  "Ttru", "Hippo"
)

# Create a vector for language complexity
log10_brain_mass = c(
   2.85,
  3.43, 3.33, 3.82, 3.89,
  3.59, 2.94, 3.19, 2.77
)

#Make sure these are in the same order as the tip labels
m = match(my.tree$tip.label, names(my.complex)) 
my.complex = my.complex[m]

names(log10_brain_mass) = species_names
m = match(my.tree$tip.label, names(log10_brain_mass))
log10_brain_mass = log10_brain_mass[m]

# Assign discrete scores based on complexity
disc_scores = log10_brain_mass
disc_scores[which(log10_brain_mass < 2.89)] = 1  # State 1: < 2.89
disc_scores[which(log10_brain_mass >= 2.89 & log10_brain_mass <= 3.08)] = 2  # State 2: 2.89 - 3.08
disc_scores[which(log10_brain_mass > 3.08 & log10_brain_mass <= 3.36)] = 3  # State 3: 3.08 - 3.36
disc_scores[which(log10_brain_mass > 3.36)] = 4  # State 4: > 3.36

Q = matrix(data=c(
  0,  2,  4,  6,    
  2,  0,  2,  4,    
  4,  2,  0,  2,       
  6,  4,  2,  0
), nrow=4, byrow=TRUE, dimnames=list(c("1", "2", "3", "4"), c("1", "2", "3", "4")))

disc_scores = disc_scores[m]

# Check if the names match the tip labels in the tree
tree_tip_labels = my.tree$tip.label

# Ensure numeric states for ASR
numeric_states = as.numeric(as.factor(disc_scores))

names(log10_brain_mass) = my.tree$tip.label

# Reassign disc_scores with correct matching
disc_scores = ifelse(
    log10_brain_mass < 2.89, 1,
    ifelse(log10_brain_mass <= 3.08, 2,
           ifelse(log10_brain_mass <= 3.36, 3, 4)
    )
)

# Verify names match
names(disc_scores) = names(log10_brain_mass)

mp_custom = asr_max_parsimony(tree=my.tree, tip_states=numeric_states, Nstates=4, transition_costs=Q)


pie.matrix = matrix(mp_custom$ancestral_likelihoods, ncol=4)
rownames(pie.matrix) = seq(10,17) #(number of taxa+1, number of taxa+number of rows)
colnames(pie.matrix) = c("1", "2", "3", "4")

plotTree(my.tree, offset=0.5, main="Ancestral Reconstruction with Custom Transition Costs")

tiplabels(pie = to.matrix(disc_scores, sort(unique(disc_scores))),
          piecol=c("yellow", "hotpink", "seagreen", "navy"),
          cex=0.3)

nodelabels(node = as.numeric(rownames(pie.matrix)),
           pie = pie.matrix, piecol = c("yellow", "hotpink", "seagreen", "navy"),
           cex = 0.5)

legend("bottomright",
       legend = c("< 2.89", "2.89 - 3.08", "3.08 - 3.36", "> 3.36"),
       fill = c("yellow", "hotpink", "seagreen", "navy"),
       bty = "n", title = "Brain Size")
```
